### 2024-01-14 nondirectional_ms_2024.Rmd
Change 2023 to 2024 + add exp name to the datasets analysis loop + RMD adjustments + ignore included additional datasets + updated figures + added appendix A and changed appndix names to B-F. Not including result files (big files)

author: 
  - name          : "Itay Yaron"
    affiliation   : "1"
    corresponding : yes
    address       : "Haim Levanon 55, Tel Aviv, Israel"
    email         : "mufc.itay@gmail.com"
  - name          : "Nathan Faivre"
    affiliation   : "2"
  - name          : "Liad Mudrik"
    affiliation   : "1,3"
  - name          : "Matan Mazor"
    affiliation   : "4,5"

affiliation:
  - id            : "1"
    institution   : "Sagol School of Neuroscience, Tel Aviv University"
  - id            : "2"
    institution   : "University Grenoble Alpes, University Savoie Mont Blanc, CNRS, LPNC"
  - id            : "3"
    institution   : "School of Psychological Sciences, Tel Aviv University"
  - id            : "4"
    institution   : "Department of Psychological Sciences, Birkbeck, University of London"
  - id            : "5"
    institution   : "Wellcome Centre for Human Neuroimaging, UCL"

authornote: |
  All datasets and analysis scripts used in this manuscript are publicly available online: https://github.com/mufcItay/NDT.

@@

@@

library(groundhog)
groundhog.library(c(
  'ggplot2',
  'stringr',
  'gridExtra',
  'grid',
  'ggh4x',
  'foreach',
  'ggtext',
  'data.table',
  'ggridges',
  'ggpubr',
  'scales',
  'dplyr',
  'tidyr',
  'pracma',
  'extraDistr',
  'patchwork',
  'HDInterval',
  'svglite',
  'MCMCpack',
  'BayesFactor',
  'nleqslv',
  'xfun'), '2023-07-22')
# separate github libraries
# groundhog.library(c(
#   'github::mufcItay/signcon'), '2023-09-09')
library(signcon)

@@

# the analysis and simulations for appendix B.
# to run the analysis again, uncomment the (commented) run_appendixB line and comment the read.csv line which reads the results of an existing simulations run.
source('appendix\\appendix_B.R')
# appendixA_res <- run_appendixB(apndxB_conf_FAs, apndxB_conf_sensitivity)
appendixA_res <- read.csv('appendix\\results\\2023-12-13_21_01_53Appendix_A_QUID_OANOVA.csv')

# the analysis and simulations for appendix C.
# to run the analysis again, uncomment the (commented) run_appendixC line and comment the read.csv line which reads the results of an existing simulations run (keeping the save_plot_appendixC line.)
source('appendix\\appendix_B.R')
# l
# appendixC_res <- run_appendixC(apndx_b_conf)
appendixC_res <- read.csv('appendix\\results\\2023-12-06_22_35_58Appendix_B.csv')
save_plot_appendixB(appendixC_res)

@@

# the analysis and simulations for appendix F.
# to run the analysis again, uncomment the (commented) run_appendixF line and comment the read.csv line which reads the results of an existing simulations run.
source('appendix\\appendix_E.R')
apndxE_N_p <- N_p
apndxE_N_t <- N_t
# appendixF_res <- run_appendixF(apndx_e_conf)
appendixF_res <- read.csv('appendix\\results\\2023-12-09_11_21_53Appendix_E.csv')
apndx_F_median_cnd_power <- appendixF_res %>%
  filter(N_p == median(appendixF_res$N_p), 
         N_t == median(appendixF_res$N_t),
         sds_ratio == median(appendixF_res$sds_ratio)) %>%
  pull(Power)
```

@@

On the face of it, pronounced individual differences in unconscious processing effects on cognition and perception seem possible, even likely. Indeed, previous investigations revealed heterogeneity in susceptibility to the attentional blink [@martens2006a], in the effects of stimulus onset asynchrony (SOA) on metacontrast masking [@albrecht2010a], the speed of prioritizing meanignful stimuli over noise [@sklar2021a] in both breaking continuous flash suppression (b-CFS; [@jiang2007a] and breaking repeated mask suppression tasks (b-RMS; [@abir2020a], and in the effects of visual imagery on conscious perception in a binocular rivalry setting [@dijkstra2019a]. Some qualitative differences have been linked to variability in processing speed [@martens2006a], genetics [@maksimov2013a], and brain anatomy and physiology [@gaal2011a; @boy2010a]. Critically, in other behavioural paradigms, unconscious stimuli had opposite effects on different participants. Bolger and colleagues [-@bolger2019a] showed that while most participants responded faster to upright faces in a b-CFS task, some responded faster to upside-down faces. Other findings from several different study groups have repeatedly revealed that masked priming effects changed in magnitude and even flipped in sign as a function of the interval between prime and target [@boy2014a; @boy2010b; @parkinson2014a; @schlaghecken2004a]. Moreover, different hypotheses were laid out over the years regarding the driving mechanisms of the counter-intuitive 'negative' priming effects. Among others, response inhibition of initial prime activations [@eimer2003a; @boy2010c], or neural habituation [@jacob2021a] were suggested. Taken together, it is not clear if, and to what extent, unconscious effects are subject to meaningful individual variability. Crucially, if they are, then some previously reported null results might actually be true effects, masked by such variability.

The paper proceeds as follows. We first simulate a setting where a strong effect of unconscious processing on behaviour is entirely missed in standard analysis, due to pronounced inter-individual differences. We then show that the same effect is revealed when using three tests that are robust to population variability: the global null prevalence test [@donhauser2018a], Bayesian hierarchical modelling [@haaf2019a], and a test based on analysis of variance (ANOVA; [@miller2018a]. Importantly, unlike common measures of reliability which are used to directly estimate individual differences (see @parsons2019a), the above tests do not quantify individual differences, but *measure group effects in a way that is robust to such differences*. Hence, they provide researchers with the appropriate tools for detecting unconscious effects in face of potential individual differences, without depending on that being the case.

@@

To provide a conceptual demonstration of how true causal effects of unconscious processing can be masked by inter-individual differences in effect signs, we simulated a typical experiment using a within-participants manipulation (Figure 1). Specifically, we generated trial-by-trial data from a standard unconscious priming experiment. For each simulated participant, we generated reaction time data from two conditions (corresponding to congruent and incongruent primes in unconscious processing studies). Individual-level effect sizes (in milliseconds) were sampled from a normal distribution centred at zero ($e_{i} \sim {\mathcal{N}}(0, \sigma_{b})$, where $e_{i}$ denotes the true effect size of the $i^{th}$ participant and $\sigma_b$ the between-participant standard deviation. Then, the trial-by-trial reaction times (RTs) of each participant and condition were generated according to each participant's true effect score ($e_{i}$), the relevant condition ($c \in \{1, 0\}$, where $c=1$ denotes the incongruent condition, and $c=0$ denotes the congruent condition), and the within-participant standard deviation ($\sigma_{w}$) ($RT_{i,c} \sim {\mathcal{N}} (0, \sigma_{w}) +c*e_{i}$).^[Similar results were obtained with more realistic Wald distributions as detailed in Appendix A.]

@@

To reiterate, a standard t-test misses existing individual-level effects because, operating on individual-level summary statistics, it is oblivious to within-participant variability in the dependent variable. In recent years, researchers sought to address this limitation, advocating for the use of statistical methods that incorporate both within and between-participant variability. Specifically, three approaches were proposed. First, the *prevalence global null* approach (@donhauser2018a; henceforth GNT) tests if the prevalence of individual-level effects in a given population (the proportion of individuals showing an effect) is greater than zero. The prevalence approach relies on a two stages procedure. In the first stage, effects are tested at the individual level using a standard hypothesis-testing approach. In the second stage, the proportion of observed individual-level effects is tested against the alpha level, or type-1 error rate, using a binomial test. A significantly higher prevalence means the *global null* hypothesis, according to which individual shows true effect, can be rejected. Second, the *qualitative individual differences* approach (@rouder2020a; @rouder2021a; @haaf2019a; henceforth QUID) quantifies the relative support for the presence of “qualitative differences” in effects, that is, inter-individual differences in effect signs, by performing a Bayesian model comparison over a family of hierarchical models with different constraints [@haaf2019a]. Third, Miller & Schwarz [-@miller2018a] introduce a parametric and frequentist test, based on ANOVA. Specifically, their Omnibus ANOVA test (henceforth OANOVA) probes the joint null hypothesis that there are no systematic differences neither between experimental conditions across individuals, nor within individuals and across trials. Together, this is equivalent to the *global null* scenario we presented above.

We applied the tests to our simulated data. For QUID, we used the default priors from the original publication of QUID [@rouder2021a]. For GNT and OANOVA, we used an $\alpha$ of `r oanova_alpha` to examine individual-level and group-level effects. For QUID, we considered $BF>`r bf_criteria`$ as evidence for an effect, $BF< \frac{1}{`r bf_criteria`}$ as evidence for no effect (*global null*), and values between these thresholds ($\frac{1}{`r bf_criteria`}\leq BF\leq `r bf_criteria`$) as inconclusive [@jeffreys1998a]. Reassuringly, all tests were able to differentiate between the two simulated scenarios, providing very strong evidence for an effect in the *non-directional differences* scenario, but not in the *global null* one. Specifically, According to GNT, the prevalence of effects on RT was clearly above zero in the *non-directional differences* simulation (using a two-sided t-test for the individual-level test; `r round(100 * nde_gnt_res$stat,2)`% of significant effects, one-sided $CI_{95}$ = [`r round(100 * nde_gnt_res$ci_low,2)` `r round(100 * nde_gnt_res$ci_high,2)`], p `r apa_p(nde_gnt_res$p)`), but this proportion was not reliably higher than $\alpha$ in the *global null* simulation (`r round(100 * sn_gnt_res$stat,2)`% of significant effects, one-sided $CI_{95}$ = [`r round(100 * sn_gnt_res$ci_low,2)` `r round(100 * sn_gnt_res$ci_high,2)`], p `r apa_p(sn_gnt_res$p)`). Using the QUID method, a random effects model with individual-level effects was overwhelmingly preferred in the *non-directional differences* simulation ($BF$= `r formatC(nde_bf, format = 'e', 2)`), but a null model was preferred in the *global null* simulation ($BF$= `r round(sn_bf,2)`). Similarly, the OANOVA test revealed significant results in the *non-directional differences* scenario (F(`r nde_OANOVA_res$df_num`, `r nde_OANOVA_res$df_denom`) = `r round(nde_OANOVA_res$F,2)`, p `r paste0(ifelse(nde_OANOVA_res$p >= .001, '=', ''), apa_p(nde_OANOVA_res$p))`), and a non-significant effect in the *global null* simulation (F(`r sn_OANOVA_res$df_num`, `r sn_OANOVA_res$df_denom`) = `r round(sn_OANOVA_res$F,2)`, p `r paste0(ifelse(sn_OANOVA_res$p >= .001, '=', ''),apa_p(sn_OANOVA_res$p))`).

@@

To that end, the effects of interest were tested using GNT, QUID, and the OANOVA tests (see Supplementary Figure 1 for an analysis of the significant directional effects which were excluded). GNT was applied to all `r n_empiric_ns` effects. In contrast, QUID and OANOVA were used on subsets of `r n_empiric_ns_quid` and `r n_empiric_ns_oanova` of these effects, respectively (omitting five effects of signal detection sensitivity, d’, from both tests, and one additional RT interaction from the QUID analysis, as its current implementation only supports simple RT effects).
All tests agreed on finding no reliable evidence for non-directional unconscious effects. According to GNT, the prevalence statistic was zero in `r round(100*p_gnt_stat_zero,2)`% of the effects (maximal observed prevalence = `r ns_gnt_res %>% filter(is_sim == FALSE) %>% mutate(prev=round(100*gnt.stat, 2)) %>% pull(prev) %>% max()`%; see Fig. 2A), and the 95% one-sided CI included $\alpha=5$% in all of them (hence, for all effects the prevalence of effects did not exceed the expected rate under the global null hypothesis). Similarly, for both QUID and the OANOVA tests, no single $BF$ or p-value revealed evidence for an effect (maximal $BF_{10}$ = `r round(ns_quid_res %>% filter(is_sim == FALSE) %>% pull(quid_bf) %>% max(), 2)` and all p-values > `r oanova_alpha`; see Figure 2B, C). Notably, QUID obtained moderate evidence for the *global null* model in `r round(100*p_null_QUID,2)`% of the cases (see Fig. 2B). The remaining effects were inconclusive. Hence, for the effects collected here, in the case of unconscious processing, the three tests revealed a highly similar pattern of results, consistent with a strong interpretation of previously reported null results as revealing the genuine absence of a causal effect of unconsciously perceived stimuli on behaviour.

@@

In the next section, we introduce a novel non-directional test that takes into account population heterogeneity to infer group-level effects. The test is both frequentist and non-parametric, which addresses the above issues. Similarly to the OANOVA test it promises a control for long term error-rates, but unlike it, our test does not assume a parametric model of the data-generating process. Using a continuous within-participant summary statistic, it is also more statistically powerful than approaches that focus on a dichotomous notion of effect prevalence (see Appendix C).

@@

Our test assumes that a within-participant effect is convincing if it is consistently evident across different trials. To test this, we can split the trials of an individual into two random halves, and ask whether both halves show the same qualitative effect (e.g., for both halves, the performance in the congruent condition is higher than in the incongruent condition; see Fig. 3A). By doing this many times, we can measure how often the two halves agree. Following this strategy, we estimate the consistency of effect signs within each individual by measuring the frequency of consistent results across splits. Then, we compare the group-mean consistency score against a null distribution: 10,000 samples of group-level consistency scores, obtained after randomly shuffling the experimental condition labels within participants (here, to speed up the computational process, for each participant 100 permutations were created, from which we randomly sampled a single permutation in each null distribution sample; @stelzer2013a). Hence, our null distribution reflects the expected consistency of within-participant effects when the dependent variable of no single participant is sensitive to the experimental manipulation. An easy-to-use implementation of the sign consistency test is available as part of the signcon R package ([https://github.com/mufcItay/signcon](https://github.com/mufcItay/signcon); see Appendix D for extensions of the test to use cases that diverge from simple mean difference between conditions).

@@

test_full_prevalence <- binom.test(sc_sig_n, n_empiric_ns, full_prevalence_exp_rate, alternative = 'less')
test_half_prevalence <- binom.test(sc_sig_n, n_empiric_ns, half_prevalence_exp_rate, alternative = 'less')
```
Together, four different analysis methods support the conclusion that by and large, unconscious priming effects are not masked by individual differences. Yet one can still claim that these statistical tests are simply not sensitive enough to detect qualitatively variable, non-directional effects, even when those exist. To test this claim, we conducted two additional analyses: First, we used simulations to estimate the sensitivity of our solution to non-directional effects with various effect sizes, determined according to previous analyses on unconscious processing [@meyen2022a] and cognitive control [@rouder2023a]. The results corroborated the concerns for lack of power when using common unconscious processing settings of the number of participants and trials (see @baker2021 for a detailed analysis of the contribution of both factors to power). Yet, we conducted further analysis showing that given the (low) power estimate we found, the number of significant effects obtained by the sign consistency test would be surprisingly low if an effect existed in all or even half of the datasets (see Appendix F). Second, to provide positive-control for these methods, and show that they can be used to reveal such hidden effects in other fields, we collected additional, openly accessible, datasets from studies conducted in different fields of research within experimental psychology. We then used our non-parametric test on these datasets, demonstrating its potential benefit in determining whether a null result at the group level hides true, but variable, effects at the individual participant level.

@@

# get effects for which non-directional effects were already reported
machado_etal_fld <- 'datasets\\additional_analyses\\Machado_etal_2007_2009'
source(paste(machado_etal_fld, "process_Machado_2007_2009.R", sep = .Platform$file.sep))
machado_2007_res <- analyze_machado(2007, machado_etal_fld)
machado_2009_res <- analyze_machado(2009, machado_etal_fld)
machado_2007_350 <- machado_2007_res[machado_2007_res$SOA == 350,]
machado_2009_650 <- machado_2009_res[machado_2009_res$SOA == 650,]
```

@@

Importantly, although we cannot conclusively attribute these non-directional effects to social setting versus condition order in the first example, or to response versus stimulus in the second, they both constitute examples of true effects that were masked by inter-individual differences. The absence of a directional effect in Battich et al is indicated by the fact that on average, participants showed similar levels of multisensory integration in the first and second parts of all three experiments showing non-directional sign consistency effects (`r apa_print(res_battich_resp1_t)$full_result`, `r apa_print(res_battich_resp2_t)$full_result`, and `r apa_print(res_battich_rt2_t)$full_result`). In the case of confidence effects, response mapping was not counterbalanced across participants in many of the considered datasets. This way, the absence of a directional effect of response is also indicative of the absence of a directional effect of stimulus. Together, these previously hidden non-directional findings make the absence of significant non-directional effects in unconscious processing a more convincing indication of the true absence of such effects at the individual-participant level.

@@

It is important to note that our claim here is not about the presence of individual differences in unconscious processing in general, but about the likelihood that such differences in effect signs may be responsible for null group-level findings. Indeed, previous studies revealed inter-individual differences in the magnitude of unconscious processing effects [@gaal2011a; @cohen2009a; @boy2010a]. For example, Van Gaal et al., [-@gaal2011a] used fMRI and a meta-contrast masked arrows-priming task, to show that grey matter density is correlated with the size of unconscious motor priming effects. Yet importantly, in this experiment effects were defined according to the assumption that *incongruent* trials are performed slower than *congruent* trials (trials in which primes and targets pointed to opposing and the same direction, respectively. This assumption of group coherence in effect signs is prevalent in consciousness science, and in cognitive science more broadly, with few exceptions (for example, see Bolger et al., [-@bolger2019a] for a study where the direction of face orientation effects was not assumed in advance). Here, in contrast, we asked whether relaxing the assumption of effect sign uniformity could reveal unconscious effects that remain undetected using standard directional approaches.

>>>>

author: 
  - name          : "Itay Yaron"
    affiliation   : "1"
    corresponding : yes
    address       : "Haim Levanon 55, Tel Aviv, Israel"
    email         : "mufc.itay@gmail.com"
  - name          : "Nathan Faivre"
    affiliation   : "2"
  - name          : "Liad Mudrik"
    affiliation   : "1,3"
  - name          : "Matan Mazor"
    affiliation   : "4,5,6"

affiliation:
  - id            : "1"
    institution   : "Sagol School of Neuroscience, Tel Aviv University"
  - id            : "2"
    institution   : "University Grenoble Alpes, University Savoie Mont Blanc, CNRS, LPNC"
  - id            : "3"
    institution   : "School of Psychological Sciences, Tel Aviv University"
  - id            : "4"
    institution   : "Department of Psychological Sciences, Birkbeck, University of London"
  - id            : "5"
    institution   : "Wellcome Centre for Human Neuroimaging, UCL"
  - id            : "6"
    institution   : "All Souls College and Department of Experimental Psychology, University of Oxford"

authornote: |
  All data, and analysis code are available at [https://github.com/mufcItay/NDT]. This study's analysis was not pre-registered.

  MM was funded by a post-doctoral research fellowships from All Souls College at the University of Oxford.

@@

  Significance statement: In consciousness science, negative results play an important theoretical role, revealing which cognitive and perceptual functions cannot take place without consciousness. We challenged this strong, traditional, interpretation of null results, by a never-tested weaker interpretation: participants may have been affected by unconscious manipulations, yet in opposing directions, resulting in non-significant results at the group-level. To ask this question, we used three existing statistical tests, and developed a novel non-parametric test tailored to arbitrate between the two competing interpretations. By and large, our results provide support for the traditional, strong interpretation: consciousness is necessary for the probed functions. Importantly, in contrast to the results obtained for unconscious effects, we found effects that were hidden by individual-differences in various other fields. To facilitate the adoption of our new test, we offer best-practice recommendations for testing hypotheses without assuming consistent effect signs, and publish our novel test as a ready-to-use R package.

@@

library(groundhog)
groundhog.library(c(
  'ggplot2',
  'stringr',
  'gridExtra',
  'ggh4x',
  'foreach',
  'ggtext',
  'ggridges',
  'ggpubr',
  'scales',
  'dplyr',
  'tidyr',
  'pracma',
  'extraDistr',
  'patchwork',
  'MCMCpack',
  'BayesFactor',
  'xfun'), '2023-07-22')
# from github (Commit: 0e04c3c)
library(signcon)

@@

# we run the analysis of appendix A right before presenting the appendix (see below).

# the analysis and simulations for appendix B.
# to run the analysis again, uncomment the (commented) run_appendixB line and comment the read.csv line which reads the results of an existing simulations run.
source('appendix\\appendix_B.R')
# appendixB_res <- run_appendixB(apndxB_conf_FAs, apndxB_conf_sensitivity)
appendixB_res <- read.csv('appendix\\results\\2024-01-07_23_39_16Appendix_B_QUID_OANOVA.csv')

# the analysis and simulations for appendix C.
# to run the analysis again, uncomment the (commented) run_appendixC line and comment the read.csv line which reads the results of an existing simulations run (keeping the save_plot_appendixC line.)
source('appendix\\appendix_C.R')
# appendixC_res <- run_appendixC(apndx_b_conf)
appendixC_res <- read.csv('appendix\\results\\2023-12-29_18_48_11Appendix_C.csv')
save_plot_appendixC(appendixC_res)

@@

# the analysis and simulations for appendix F.
# to run the analysis again, uncomment the (commented) run_appendixF line and comment the read.csv line which reads the results of an existing simulations run.
source('appendix\\appendix_F.R')
apndxF_N_p <- N_p
apndxF_N_t <- N_t
# appendixF_res <- run_appendixF(apndx_e_conf)
appendixF_res <- read.csv('appendix\\results\\2024-01-09_16_46_09Appendix_F.csv')
apndx_F_median_cnd_power <- appendixF_res %>%
  filter(N_p == median(appendixF_res$N_p), 
         N_t == median(appendixF_res$N_t),
         sds_ratio == median(appendixF_res$sds_ratio)) %>%
  pull(Power)
```

@@

On the face of it, pronounced individual differences in unconscious processing effects on cognition and perception seem possible, even likely. Indeed, previous investigations revealed heterogeneity in susceptibility to the attentional blink [@martens2006a], in the effects of stimulus onset asynchrony (SOA) on metacontrast masking [@albrecht2010a], the speed of breaking perceptual suppression [@sklar2021a], both in the breaking continuous flash suppression paradigm (b-CFS; [@jiang2007a] and the breaking repeated mask suppression one (b-RMS; [@abir2020a], and in the effects of visual imagery on conscious perception in a binocular rivalry setting [@dijkstra2019a]. Some qualitative differences have been linked to variability in processing speed [@martens2006a], genetics [@maksimov2013a], and brain anatomy and physiology [@gaal2011a; @boy2010a]. Critically, in other behavioural paradigms, unconscious stimuli had opposite effects on different participants. Bolger and colleagues [-@bolger2019a] showed that while most participants responded faster to upright faces in a b-CFS task, some responded faster to upside-down faces. Other findings from several different study groups have repeatedly revealed that masked priming effects changed in magnitude and even flipped in sign as a function of the interval between prime and target [@boy2014a; @boy2010b; @parkinson2014a; @schlaghecken2004a]. Moreover, different hypotheses were laid out over the years regarding the driving mechanisms of the counter-intuitive 'negative' priming effects. Among others, response inhibition of initial prime activations [@eimer2003a; @boy2010c], or neural habituation [@jacob2021a] were suggested. Taken together, it is not clear if, and to what extent, unconscious effects are subject to meaningful individual variability. Crucially, if they are, then some previously reported null results might actually be true effects, masked by such variability.

The paper proceeds as follows. We first simulate a setting where a strong effect of unconscious processing on behaviour is entirely missed in standard analysis, due to pronounced inter-individual differences. We then show that the same effect is revealed when using three tests that are robust to population variability: the global null prevalence test [@donhauser2018a], Bayesian hierarchical modelling [@haaf2019a], and a test based on analysis of variance (ANOVA; [@miller2018a]. Importantly, unlike common measures of reliability which are used to directly estimate individual differences (see @parsons2019a), the above tests do not quantify individual differences, but *measure group effects in a way that is robust to such differences*. Hence, they provide researchers with the appropriate tools for detecting unconscious effects even if pronounced individual differences exist, without depending on that being the case.

@@

To provide a conceptual demonstration of how true causal effects of unconscious processing can be masked by inter-individual differences in effect signs, we simulated a typical experiment using a within-participants manipulation (Figure 1). Specifically, we generated trial-by-trial data from a standard unconscious priming experiment. For each simulated participant, we generated reaction time data from two conditions (corresponding to congruent and incongruent primes in unconscious processing studies). Individual-level effect sizes (in milliseconds) were sampled from a normal distribution centred at zero ($e_{i} \sim {\mathcal{N}}(0, \sigma_{b})$, where $e_{i}$ denotes the true effect size of the $i^{th}$ participant and $\sigma_b$ the between-participant standard deviation. Then, the trial-by-trial reaction times (RTs) of each participant and condition were generated according to each participant's true effect score ($e_{i}$), the relevant condition ($c \in \{1, 0\}$, where $c=1$ denotes the incongruent condition, and $c=0$ denotes the congruent condition), and the within-participant standard deviation ($\sigma_{w}$) ($RT_{i,c} \sim {\mathcal{N}} (0, \sigma_{w}) +c*e_{i}$).^[Similar results were obtained with more realistic Wald RT distributions as detailed in Appendix A.]

@@

To reiterate, a standard t-test misses existing individual-level effects because, operating on individual-level summary statistics, it is oblivious to within-participant variability in the dependent variable. In recent years, researchers sought to address this limitation, advocating for the use of statistical methods that incorporate both within and between-participant variability. Specifically, three approaches were proposed. First, the *prevalence global null* approach (@donhauser2018a; henceforth GNT) tests if the prevalence of individual-level effects in a given population (the proportion of individuals showing an effect) is greater than zero. The prevalence approach relies on a two stages procedure. In the first stage, effects are tested at the individual level using a standard hypothesis-testing approach. In the second stage, the proportion of observed individual-level effects is tested against the alpha level, or type-1 error rate, using a binomial test. A significantly higher prevalence means the *global null* hypothesis, according to which no individual shows a true effect, can be rejected. Second, the *qualitative individual differences* approach (@rouder2020a; @rouder2021a; @haaf2019a; henceforth QUID) quantifies the relative support for the presence of “qualitative differences” in effects, that is, inter-individual differences in effect signs, by performing a Bayesian model comparison over a family of hierarchical models with different constraints [@haaf2019a]. Third, Miller & Schwarz [-@miller2018a] introduce a parametric and frequentist test, based on ANOVA. Specifically, their Omnibus ANOVA test (henceforth OANOVA) probes the joint null hypothesis that there are no systematic differences neither between experimental conditions across individuals, nor within individuals and across trials. Together, this is equivalent to the *global null* scenario we presented above.

We applied the tests to our simulated data. For QUID, we used the default priors from the original publication [@rouder2021a]. For GNT and OANOVA, we used an $\alpha$ of `r oanova_alpha` to examine individual-level and group-level effects. For QUID, we considered $BF>`r bf_criteria`$ as evidence for an effect, $BF< \frac{1}{`r bf_criteria`}$ as evidence for no effect (*global null*), and values between these thresholds ($\frac{1}{`r bf_criteria`}\leq BF\leq `r bf_criteria`$) as inconclusive [@jeffreys1998a]. Reassuringly, all tests were able to differentiate between the two simulated scenarios, providing very strong evidence for an effect in the *non-directional differences* scenario, but not in the *global null* one. Specifically, According to GNT, the prevalence of effects on RT was clearly above zero in the *non-directional differences* simulation (using a two-sided t-test for the individual-level test; `r round(100 * nde_gnt_res$stat,2)`% of significant effects, one-sided $CI_{95}$ = [`r round(100 * nde_gnt_res$ci_low,2)` `r round(100 * nde_gnt_res$ci_high,2)`], p `r apa_p(nde_gnt_res$p)`), but this proportion was not higher than $\alpha$ in the *global null* simulation (`r round(100 * sn_gnt_res$stat,2)`% of significant effects, one-sided $CI_{95}$ = [`r round(100 * sn_gnt_res$ci_low,2)` `r round(100 * sn_gnt_res$ci_high,2)`], p =`r apa_p(sn_gnt_res$p)`). Using the QUID method, a random effects model with individual-level effects was overwhelmingly preferred in the *non-directional differences* simulation ($BF$= `r formatC(nde_bf, format = 'e', 2)`), but a null model was preferred in the *global null* simulation ($BF$= `r round(sn_bf,2)`). Similarly, the OANOVA test revealed significant results in the *non-directional differences* scenario (F(`r nde_OANOVA_res$df_num`, `r nde_OANOVA_res$df_denom`) = `r round(nde_OANOVA_res$F,2)`, p `r paste0(ifelse(nde_OANOVA_res$p >= .001, '=', ''), apa_p(nde_OANOVA_res$p))`), and a non-significant effect in the *global null* simulation (F(`r sn_OANOVA_res$df_num`, `r sn_OANOVA_res$df_denom`) = `r round(sn_OANOVA_res$F,2)`, p `r paste0(ifelse(sn_OANOVA_res$p >= .001, '=', ''),apa_p(sn_OANOVA_res$p))`).

@@

To that end, the effects of interest were tested using GNT, QUID, and the OANOVA tests (see Supplementary Figure 1 for an analysis of the significant directional effects which were excluded). GNT was applied to all `r n_empiric_ns` effects. In contrast, QUID and OANOVA were used on subsets of `r n_empiric_ns_quid` and `r n_empiric_ns_oanova` of these effects, respectively (omitting five effects of signal detection sensitivity, d’, from both tests, and one additional RT interaction from the QUID analysis, as its current implementation only supports simple RT effects).
All tests agreed on finding no reliable evidence for non-directional unconscious effects. According to GNT, the prevalence statistic was zero in `r round(100*p_gnt_stat_zero,2)`% of the effects (maximal observed prevalence = `r ns_gnt_res %>% filter(is_sim == FALSE) %>% mutate(prev=round(100*gnt.stat, 2)) %>% pull(prev) %>% max()`%; see Fig. 2A), and the 95% one-sided CI included $\alpha=5$% in all of them. Hence, for all effects the prevalence of effects did not exceed the expected rate under the global null hypothesis. Similarly, for both QUID and the OANOVA tests, no single $BF$ or p-value revealed evidence for an effect (maximal $BF_{10}$ = `r round(ns_quid_res %>% filter(is_sim == FALSE) %>% pull(quid_bf) %>% max(), 2)` and all p-values > `r oanova_alpha`; see Figure 2B, C). Notably, QUID obtained moderate evidence for the *global null* model in `r round(100*p_null_QUID,2)`% of the cases (see Fig. 2B). The remaining effects were inconclusive. Hence, for the effects collected here, in the case of unconscious processing, the three tests revealed a highly similar pattern of results, consistent with a strong interpretation of previously reported null results as revealing the genuine absence of a causal effect of unconsciously perceived stimuli on behaviour.

@@

In the next section, we introduce a novel non-directional test that takes into account population heterogeneity to infer group-level effects. The test is both frequentist and non-parametric, which addresses the above issues. Similarly to the OANOVA test, it promises a control for long term error-rates, but unlike it, our test does not assume a parametric model of the data-generating process. Using a continuous within-participant summary statistic, it is also more statistically powerful than approaches that focus on a dichotomous notion of effect prevalence (see Appendix C).

@@

Our test assumes that a within-participant effect is convincing if it is consistently evident across different trials. To test this, we can split the trials of an individual into two random halves, and ask whether both halves show the same qualitative effect (e.g., for both halves, the performance in the congruent condition is higher than in the incongruent condition; see Fig. 3A). By doing this many times, we can measure how often the two halves agree. Following this strategy, we estimate the consistency of effect signs within each individual by measuring the frequency of consistent results across splits. Then, we compare the group-mean consistency score against a null distribution: 10,000 samples of group-level consistency scores, obtained after randomly shuffling the experimental condition labels within participants (here, to speed up the computational process, for each participant, 100 permutations were created, from which we randomly sampled a single permutation in each null distribution sample; @stelzer2013a). Hence, our null distribution reflects the expected consistency of within-participant effects when the dependent variable of no single participant is sensitive to the experimental manipulation. An easy-to-use implementation of the sign consistency test is available as part of the signcon R package ([https://github.com/mufcItay/signcon](https://github.com/mufcItay/signcon); see Appendix D for extensions of the test to use cases that diverge from simple mean difference between conditions).

@@

test_full_prevalence <- binom.test(sc_sig_n, n_empiric_ns, full_prevalence_exp_rate, alternative = 'less')
test_half_prevalence <- binom.test(sc_sig_n, n_empiric_ns, half_prevalence_exp_rate, alternative = 'less')
```
Together, four different analysis methods support the conclusion that by and large, unconscious priming effects are not masked by individual differences. Yet one can still claim that these statistical tests are simply not sensitive enough to detect qualitatively variable, non-directional effects, even when those exist. To test this claim, we conducted two additional analyses: First, we used simulations to estimate the sensitivity of our solution to non-directional effects with various effect sizes, determined according to previous analyses on unconscious processing [@meyen2022a] and cognitive control [@rouder2023a]. The results corroborated the concerns for lack of power when using common unconscious processing settings of the number of participants and trials (see @baker2021 for a detailed analysis of the contribution of both factors to power). Yet, we conducted further analysis showing that given the (low) power estimate we found, the number of significant effects obtained by the sign consistency test would be surprisingly low if an effect existed in all or even half of the datasets (see Appendix F). Second, to provide positive-control for these methods and show that they can be used to reveal such hidden effects in other fields, we collected additional, openly accessible, datasets from studies conducted in different fields of research within experimental psychology. We then used our non-parametric test on these datasets, demonstrating its potential benefit in determining whether a null result at the group level hides true, but variable, effects at the individual participant level.

@@

# get effects for which non-directional effects were already reported
machado_etal_fld <- 'datasets\\additional_analyses\\Machado_etal_2007_2009'
# the data was shared with us privately, so here we return the obtained statistics if data is not available
source(paste(machado_etal_fld, "process_Machado_2007_2009.R", sep = .Platform$file.sep))
machado_2007_res <- analyze_machado(2007, machado_etal_fld)
machado_2009_res <- analyze_machado(2009, machado_etal_fld)
machado_2007_350 <- machado_2007_res[machado_2007_res$SOA == 350,]
machado_2009_650 <- machado_2009_res[machado_2009_res$SOA == 650,]
```

@@

Importantly, although we cannot conclusively attribute these non-directional effects to social setting versus condition order in the first example, or to response versus stimulus in the second, they both constitute examples of true effects that were masked by inter-individual differences. The absence of a directional effect in Battich et al. is indicated by the fact that on average, participants showed similar levels of multisensory integration in the first and second parts of all three experiments showing non-directional sign consistency effects (`r apa_print(res_battich_resp1_t)$full_result`, `r apa_print(res_battich_resp2_t)$full_result`, and `r apa_print(res_battich_rt2_t)$full_result`). In the case of confidence effects, response mapping was not counterbalanced across participants in many of the considered datasets. This way, the absence of a directional effect of response is also indicative of the absence of a directional effect of stimulus. Together, these previously hidden non-directional findings make the absence of significant non-directional effects in unconscious processing a more convincing indication of the true absence of such effects at the individual-participant level.

@@

It is important to note that our claim here is not about the presence of individual differences in unconscious processing in general, but about the likelihood that such differences in effect signs may be responsible for null group-level findings. Indeed, previous studies revealed inter-individual differences in the magnitude of unconscious processing effects [@gaal2011a; @cohen2009a; @boy2010a]. For example, Van Gaal et al., [-@gaal2011a] used fMRI and a meta-contrast masked arrows-priming task, to show that grey matter density is correlated with the size of unconscious motor priming effects. Yet importantly, in this experiment effects were defined according to the assumption that *incongruent* trials are performed slower than *congruent* trials (trials in which primes and targets pointed to opposing and the same direction, respectively). This assumption of group coherence in effect signs is prevalent in consciousness science, and in cognitive science more broadly, with few exceptions (for example, see Bolger et al., [-@bolger2019a] for a study where the direction of face orientation effects was not assumed in advance). Here, in contrast, we asked whether relaxing the assumption of effect sign uniformity could reveal unconscious effects that remain undetected using standard directional approaches.
